Flattering Advice: Avoiding Disappointment as a Driver of Gender Discrimination

### 2024-09-27 manuscript/manuscript.Rmd
Study 1 Result (1/2)

In a large, preregistered experiment (n = 2,000), we show that advisers
consider the expectations of advisees, even when such expectations do
not provide informative insights for the underlying decision-making
problem. We find that as a result, showing expectations (but not gender)
to advisers leads men to receive more aspirational advice than women.
Incorporating expectations leads to worse advice, even when advisers are
incentivized based on the outcome of the advisee's decision. In a second
preregistered experiment (n = 700), we show that advisers respond to
incentives for interpersonal considerations. They tell people that they
are more attractive than they believe them to be, and inflate to a
greater extent when incentivized for likability rather than accuracy.
Notably, they inflate even when only incentivized for accuracy. Advisees
indeed reward such flattering advice, rating the advisers more likeable
and no less trustworthy than those who provide more accurate advice.

@@

We then recruited `r scales::comma(dim(study1_time1)[1])`
participants to serve as advisees in Stage 1 of our main experiment. These 
participants completed the same 10-item ability quiz as the earlier group, 
with the opportunity to earn 5 cents for each correct answer. After the 
quiz, we asked them to indicate their initial preference for competitors. 
Half of the participants were randomly shown the default option of the 
'High Performer Group,' highlighted in blue, while the other half were 
shown the 'Low Performer Group' as the default. This manipulation was 
designed to influence advisees' expectations, as advisees who believed 
they could compete against stronger competitors were perceived by givers 
to have higher expectations. However, participants were free to opt out 
of the default choice.

@@

```{r study1regs}
study3_advice <- study3_advice %>% dplyr::mutate(Treatment = factor(Treatment, levels = c('Low', 'High'))) %>%
  dplyr::mutate(ExpShow = ifelse(Condition == 'E',1,0))

@@

```{r study1genderdiff, fig.cap ="Advisers observed real performance and the participants' estimated performance, but not their gender. As a result of their higher expectations, men (blue) were advised to compete against the high performance group more often than women (red). The grey line shows advice absent expectations, which did not differ by gender."}
library(dplyr)
library(ggplot2)
library(magrittr)

@@

However, recall that men were more confident in their performance than
were women. Our theory thus predicts that showing expectations should
lead to more flattering advice (i.e., a recommendation to compete
against the High Performer group) for men than for women. We report a
linear probability model with advice to compete against the high group
as the outcome measure, and the advisee's gender, whether expectations
were shown to the adviser, and the interaction of the two in Column 2 of
Table \@ref(tab:study1regs). As predicted, we find a significant
interaction effect: men are more likely to be advised to compete against
the High Performers when expectations are shown than when they are
hidden ($p < 0.001$). We show this result graphically in Figure
\@ref(fig:study1genderdiff)).

@@

To examine the quality of advice, we computed the expected bonus
earnings for someone who followed the recommendations. For example, if
an adviser suggested competing against the High Performer group, we
matched the advisee against all 20 members of that group and determined
how often their score matched or exceeded that of the member. We then
multiplied this number by the respective bonus earnings (50 cents and 30
cents for the High and Low Performer group, respectively). To see if
including expectations leads to worse advise for men, we report a linear
probability model with the experimental treatment of the adviser, the
gender of the advisee, and their interaction in Column 3 of Table
\@ref(tab:study1regs). Displaying expectations led men to be advised to
compete against the High Performer group more often, and this advise
turned out to be bad: men receive worse advise than do women when
expectations are displayed, but not in their absence.[^2]

>>>>

In a large, preregistered experiment (n = 1,200), we show that advisers
consider the expectations of advisees, even when such expectations do
not provide informative insights for the underlying decision-making
problem. We find that as a result, showing expectations (but not gender)
to advisers leads men to receive more aspirational advice than women.
Incorporating expectations leads to worse advice, even when advisers are
incentivized based on the outcome of the advisee's decision. In a second
preregistered experiment (n = 700), we show that advisers respond to
incentives for interpersonal considerations. They tell people that they
are more attractive than they believe them to be, and inflate to a
greater extent when incentivized for likability rather than accuracy.
Notably, they inflate even when only incentivized for accuracy. Advisees
indeed reward such flattering advice, rating the advisers more likeable
and no less trustworthy than those who provide more accurate advice.

@@

We then recruited `r scales::comma(dim(study1_time1)[1])`
participants to serve as advisees in Stage 1 of our main experiment. These 
participants completed the same 10-item ability quiz as the earlier group, 
with the opportunity to earn 5 cents for each correct answer. After the 
quiz, we asked them to indicate their initial preference for competitors. 
Half of the participants were randomly shown the default option of the 
'High Performer Group,' highlighted in blue ("High Default" treatment), while the other half were 
shown the 'Low Performer Group' as the default ("Low Default" treatment). This manipulation was 
designed to influence advisees' expectations, as advisees who believed 
they could compete against stronger competitors were perceived by givers 
to have higher expectations. However, participants were free to opt out 
of the default choice.

We then recruited `r scales::comma(dim(study1_time2_giver)[1])`
participants for Stage 2, placing them in the role of advisers. We began
by informing them of the ability quiz that participants in the
preliminary study and Stage 1 had completed, and informed them of the
average score of all participants in the preliminary study. Advisers had
to recommend whether an advisee should compete against the Low
Performer Group or the High Performer Group (we used these terms in the survey).
We anticipated that being told to compete against High Performers is
more flattering and hence being told to compete against the Low
Performers would be disappointing if one had expected to do well.
Advisees would earn 20 cents if their score was equal to or higher than
a randomly selected member of the Low Performer Group and they chose to
compete against this group, or 50 cents if they performed as well or
better than a randomly selected member of the High Performer Group and
they chose to compete against that group. If they scored lower, they
would not receive any bonus.

Advisers were randomly assigned to one of two treatments. In the
"Performance" treatment, they only observed the score of the advisee on the
ability quiz. In the "Expectation" treatment, they observed the
score as well as the advisee's initial preference for which group to compete against.
However, since the outcome was based solely on the quiz score and advisees didn’t 
know their scores, their initial choice wasn’t useful in deciding which group had a 
higher expected bonus. Advisers gave recommendations to 10 advisees and were told that 
if their advice was used by a participant in a follow-up survey, they would receive 
the same bonus as that participant. The survey concluded with basic demographic 
questions.

Finally, we invited participants from Stage 1 back for the follow-up
survey. Following our preregistration, we kept the survey open for 7
days. In total, `r scales::comma(dim(study1_time3)[1])` participants
(`r scales::comma(dim(study1_time3[study1_time3$Gender == 'Male',])[1])`
men,
`r scales::comma(dim(study1_time3[study1_time3$Gender == 'Female',])[1])`
women) returned. The brief survey reminded them of the task they
completed in Stage 1, informed them that other participants from
Prolific had observed their real score and given them advice against
which group to compete against, and finally were reminded them of how
many questions they guessed they had answered correctly. Importantly,
they were not informed of their true score or the score of the groups
they could compete against. Participants then observed the advice from a
randomly selected adviser and made their decision.

##Results

<!-- Amanda: Wondering if we need to report the p-value of priming manipulation -->

```{r}
contingency_table <- table(study1_time1$Guess, study1_time1$default)

chi_test <- chisq.test(contingency_table)

We begin by examining the performance of the Stage 1 participants.
Because our treatment takes place after participants completed the
ability quiz, we would not expect a difference in performance across
the Low and High Default treatments. Indeed, the two groups scored
no different from one another
(`r round(mean(study1_time1[study1_time1$default == "Low",]$Score),2)`
and
`r round(mean(study1_time1[study1_time1$default == "High",]$Score),2)`
for the Low Defult and High Defult treatments, respectively,
`r papaja::apa_print(t.test(Score ~ default, data = study1_time1, var.equal = T))$statistic`).
The priming did, however, affect advisees' initial choices. In the "Low Default" treatment, 
only `r round(mean(study1_time1[study1_time1$default == "Low",]$Guess),2)*100`% 
of participants initially preferred competing against the High Performer Group, 
compared to `r round(mean(study1_time1[study1_time1$default == "High",]$Guess),2)*100`%
of participants who saw the High Performer Group highlighted ($\chi^2$(1, n = 201) = 36.62, p < .001).

```{r study1regs}

study1_advice <- study1_time2_advice %>%
  dplyr::mutate(Treatment = factor(Treatment, levels = c('Low', 'High'))) %>%
  dplyr::mutate(ExpShow = ifelse(Condition == 'E', 1, 0))

# Fit the two models
model1 <- lm(Advice ~ ExpShow * Treatment, data = study1_advice)
model2 <- lm(Advice ~ Treatment + Score, data = study1_advice %>% dplyr::filter(ExpShow == 1))

# Use sandwich package for clustered standard errors
library(sandwich)

# Clustered standard errors by GiverID
vcov_list <- list(sandwich::vcovCL(model1, cluster = ~GiverID), 
                  sandwich::vcovCL(model2, cluster = ~GiverID))

modelsummary::modelsummary(
  list(`(1)` = model1, `(2)` = model2),
  vcov = vcov_list,
  coef_map = c(
    "TreatmentHigh" = "High Default",
    "ExpShow" = "Expectation Shown",
    "ExpShow:TreatmentHigh" = "Expectation x Default",
    "Score" = "Performance",
    "(Intercept)" = "Constant"
  ),
  stars = TRUE,
  gof_map = list(list('raw' = 'nobs', 'clean' = 'N', 'fmt' = 0)),
  caption = "Column 1 displays the advice given to compete against a High Performer Group based on whether the advisees were primed with low or high default and the initial preference shown to the advisor. Column 2 displays the advice given to compete against a High Performer Group based on the default and advisees' performance when their initial preference was visible to the advisor. All standard errors are clustered at the level of the advisor."
)

However, recall that men were more confident in their performance than
were women. Our theory thus predicts that showing expectations should
lead to more flattering advice (i.e., a recommendation to compete
against the High Performer group) for men than for women. We report a
linear probability model with advice to compete against the high group
as the outcome measure, and the advisee's gender, whether expectations
were shown to the adviser, and the interaction of the two in Column 2 of
Table \@ref(tab:study3regs). As predicted, we find a significant
interaction effect: men are more likely to be advised to compete against
the High Performers when expectations are shown than when they are
hidden ($p < 0.001$). We show this result graphically in Figure
\@ref(fig:study3genderdiff)).

@@

```{r study3regs}
study3_advice <- study3_advice %>% dplyr::mutate(Treatment = factor(Treatment, levels = c('Low', 'High'))) %>%
  dplyr::mutate(ExpShow = ifelse(Condition == 'E',1,0))

@@

```{r study3genderdiff, fig.cap ="Advisers observed real performance and the participants' estimated performance, but not their gender. As a result of their higher expectations, men (blue) were advised to compete against the high performance group more often than women (red). The grey line shows advice absent expectations, which did not differ by gender."}
library(dplyr)
library(ggplot2)
library(magrittr)

@@

Next, we examine whether advisers took the advisee's expectations into
account. Column 1 of Table \@ref(tab:study3regs) reports a linear
probability model on advising to compete against the High Performer
group for advisers in the "Expectations" treatment, controlling for the
true performance of the advisee. Because each adviser made ten
recommendations, we cluster standard errors at the adviser level.
However, contrary to our expectations, we do not see a significant
effect of the expectations treatment. This may be because the induced
difference in expectations was too small.

However, recall that men were more confident in their performance than
were women. Our theory thus predicts that showing expectations should
lead to more flattering advice (i.e., a recommendation to compete
against the High Performer group) for men than for women. We report a
linear probability model with advice to compete against the high group
as the outcome measure, and the advisee's gender, whether expectations
were shown to the adviser, and the interaction of the two in Column 2 of
Table \@ref(tab:study3regs). As predicted, we find a significant
interaction effect: men are more likely to be advised to compete against
the High Performers when expectations are shown than when they are
hidden ($p < 0.001$). We show this result graphically in Figure
\@ref(fig:study3genderdiff)).

@@

To examine the quality of advice, we computed the expected bonus
earnings for someone who followed the recommendations. For example, if
an adviser suggested competing against the High Performer group, we
matched the advisee against all 20 members of that group and determined
how often their score matched or exceeded that of the member. We then
multiplied this number by the respective bonus earnings (50 cents and 30
cents for the High and Low Performer group, respectively). To see if
including expectations leads to worse advise for men, we report a linear
probability model with the experimental treatment of the adviser, the
gender of the advisee, and their interaction in Column 3 of Table
\@ref(tab:study3regs). Displaying expectations led men to be advised to
compete against the High Performer group more often, and this advise
turned out to be bad: men receive worse advise than do women when
expectations are displayed, but not in their absence.[^2]
